A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-15 podc_paper/introduction.tex
working on related work section

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues,
starting with the classic MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96}.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
Numerous papers have suggested modifications to the MS-queue \cite{list some}, but 
all still have $\Omega(p)$ worst-case step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

This problem occurs more broadly in other data structures that keep track of a set of elements, as observed by Ruppert \cite{Rup16}.
For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
have a term in their step complexity
that depends linearly on the contention $c$, i.e., 
the number of processes that access the data structure concurrently (which can be as high
as $p$ in the worst case).
Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
that amortized $\Omega(c)$ steps per operation are indeed necessary
for CAS-based implementations of lock-free bag data structures, which allow processes
to insert elements and remove an arbitrary element (chosen non-deterministically).
Since a queue trivially implements a bag, this lower bound also holds for queues.
However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
Thus, the lower bound could be stated more precisely as an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.

In this paper, we exploit this loophole to show that it is in fact possible to implement a linearizable, lock-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For simplicity of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS instructions and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use an unbounded number of CAS instructions in the worst case
and $\Omega(p)$ CAS instructions in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure, but 
leave the problem of reclaiming that memory up to a garbage collector (such as the one Java provides).

>>>>

A great deal of research in the past 25 years has been devoted to the design of linearizable, lock-free queues,
starting with the classic MS-queue of Michael and Scott \cite{DBLP:conf/podc/MichaelS96}.
Their design uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{list-some}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} dubbed this the CAS retry problem.
The same problem occurs in array-based implementations of queues \cite{list-some}.

This problem occurs more broadly in other data structures that keep track of a set of elements, as observed by Ruppert \cite{Rup16}.
For example, lock-free lists \cite{FR04,Sha15}, stacks \cite{Tre86} and search trees \cite{EFHR14} 
have a term in their step complexity
that depends linearly on the contention $c$, i.e., 
the number of processes that access the data structure concurrently (which can be as high
as $p$ in the worst case).
Attiya and Fouren \cite{DBLP:conf/opodis/AttiyaF17} proved 
that amortized $\Omega(c)$ steps per operation are indeed necessary
for CAS-based implementations of lock-free bag data structures, which allow processes
to insert elements and remove an arbitrary element (chosen non-deterministically).
Since a queue trivially implements a bag, this lower bound also holds for queues.
At first glance, this would seem to settle the question of the step complexity of lock-free queues.
However, the lower bound leaves a loophole:  it holds only if $c$ is $O(\log\log p)$.
Thus, the lower bound could be stated more precisely as an amortized bound of $\Omega(\min(c,\log\log p))$ steps per operation.

In this paper, we exploit this loophole to show that it is in fact possible to implement a linearizable, lock-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first to have step complexity that is polylogarithmic in $p$ and in $q$, the number of elements
stored in the queue.
For simplicity of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS instructions and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use an unbounded number of CAS instructions in the worst case
and $\Omega(p)$ CAS instructions in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one Java provides.

### 2022-12-15 podc_paper/related.tex
working on related work section

\section{Related Work}
In this section, we look at previous lock-free queues. The amortized
step complexity of all the lock-free queues includes an $\Omega(p)$
term that comes from the worst case, where all $p$ processes try to do
an enqueue or a dequeue simultaneously. Morrison and Afek call this
the \textit{CAS retry problem}~\cite{DBLP:conf/ppopp/MorrisonA13}.

\subsection{List-based Queues}
Michael and Scott~\cite{DBLP:conf/podc/MichaelS96} introduced a
lock-free queue which we refer to as the MS-queue. A version of it is
included in the standard Java Concurrency Package. Their idea is to
store the queue elements in a singly-linked list. A shared variable
Head points to the first node in the linked list that has not been
dequeued, and the Tail points to the last element in the queue. To
insert a node into the linked list, they use atomic primitive
operations like \texttt{LL/SC} or \texttt{CAS}. If $p$ processes try
to enqueue simultaneously, only one can succeed, and the others have
to retry. This makes the amortized number of steps $\Omega(p)$ per
enqueue. Similarly, dequeue can take $\Omega(p)$ steps.

@@

\subsection{Restricted Queues}

David introduced the first sublinear concurrent queue
\cite{DBLP:conf/wdag/David04}. Even though his algorithm does $O(1)$
steps for each operation, it is a single-enqueuer single-dequeuer
queue and uses infinite memory. 
The author states that to reduce memory usage to be bounded, the time
per operation increases linearly.

Jayanti and Petrovic introduced a wait-free poly-logarithmic
multi-enqueuer single-dequeuer
queue~\cite{DBLP:conf/fsttcs/JayantiP05}. We use their idea of having
a tournament tree among processes to agree on the linearization of
operations to design a poly-logarithmic multi-enqueuer multi-dequeuer
queue. Unlike their work, our algorithm does not put a limit on the
number of concurrent dequeuers.

>>>>

\section{Related Work}

The MS-queue \cite{DBLP:conf/podc/MichaelS96} has stood the test of time.
A version of it is
included in the standard Java Concurrency Package.  %java.util.concurrent.ConcurrentLinkedQueue
However, as mentioned above, it suffers from the CAS retry problem
which means it is lock-free but not wait-free and has an amortized step complexity
of $\Theta(p)$ per operation.

@@

\paragraph{Other Primitives and Restricted Queues}

David introduced the first sublinear-time queue
\cite{DBLP:conf/wdag/David04}, but it works only for a single enqueuer.
It uses fetch\&increment and swap primitive instructions and takes constant time per operation, but
uses unbounded memory.  A modification to use bounded space
is described, but it increases the time per operation to $\Omega(p)$.

Jayanti and Petrovic introduced a wait-free poly-logarithmic
queue~\cite{DBLP:conf/fsttcs/JayantiP05}, but it only works for a single dequeuer. 
Our implementation uses their idea of having
a tournament tree among processes to agree on the linearization of
operations.

\here{new result I found:}
Khanchandani and Wattenhofer \cite{KW18} gave a wait-free queue implementation
with $O(\sqrt{p})$ complexity using non-standard synchronization primitives
called half-increment and half-max, which can be viewed as particular kinds of
double-word read-modify-write operations.
They use this as evidence that their primitives can be more efficient than CAS
since previous CAS-based queues all required $\Omega(p)$ step complexity.
Our new implementation counters this argument.
